---
title: "2501 project"
output: html_document
date: "2025-05-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load library

```{r}
library(tidyverse)   # For data manipulation and visualization
library(lubridate)   # For handling dates
library(readr)       # For reading CSV files
library(rvest)
library(dplyr)
library(stringr)
library(ggplot2)
library(scales)  # for percent formatting if needed
```

## Load Steam Data

Project
├── project_R.Rmd <- this file
└── SteamSpyData
    ├── Action.csv
    ├── Adventure.csv
    └── ...

The data are extracted from https://steamspy.com/ on May 7th, 2025
```{r}
action_data <- read.csv("SteamSpyData/Action.csv")

adventure_data <- read.csv("SteamSpyData/Adventure.csv")

EarlyAccess_data <- read.csv("SteamSpyData/EarlyAccess.csv")

ExEarlyAccess_data <- read.csv("SteamSpyData/ExEarlyAccess.csv")

Free_data <- read.csv("SteamSpyData/Free.csv")

Indie_data <- read.csv("SteamSpyData/Indie.csv")

MMO_data <- read.csv("SteamSpyData/MMO.csv")

RPG_data <- read.csv("SteamSpyData/RPG.csv")

Simulation_data <- read.csv("SteamSpyData/Simulation.csv")

Sports_data <- read.csv("SteamSpyData/Sports.csv")

Strategy_data <- read.csv("SteamSpyData/Strategy.csv")

```

## See the data

```{r}
head(action_data)
```

X or # means the ranking of the game in that genre, we will merge them later so we will drop them.

Game means game name

Release.date is the release date of each game

Price is price of the game

Score rank(Userscore / Metascore) is the user score and metascore which is rating

Owners is the how many people own this game, as estimated

Playtime (Median) is the median playtime, first one eg 21:38 means 21 hour 38 min median total playtime, the (14:25) means typical platime oever 2 weeks is 14 hours 25 mins median.

Developer (s) is the company/person/team that made the game

Publisher (s) is the company/team that publish or distribute the game

```{r}
head(ExEarlyAccess_data)
```

HOWEVER, one of the csv, ExEarlyAccess has an extra column of Players which represent how many % of the owners plaed the game in recent period.

As ExEarlyAccess is the ONLY csv with this column, we will choose to remove it

```{r}
# Remove "Players" column from ExEarlyAccess_data only
ExEarlyAccess_data <- ExEarlyAccess_data %>% select(-Players)
```


```{r}
# Function to drop the first column (unique ID)
drop_id <- function(df) {
  df[, -1]
}

# Apply the function to each data frame
action_data <- drop_id(action_data)
adventure_data <- drop_id(adventure_data)
EarlyAccess_data <- drop_id(EarlyAccess_data)
ExEarlyAccess_data <- drop_id(ExEarlyAccess_data)
Free_data <- drop_id(Free_data)
Indie_data <- drop_id(Indie_data)
MMO_data <- drop_id(MMO_data)
RPG_data <- drop_id(RPG_data)
Simulation_data <- drop_id(Simulation_data)
Sports_data <- drop_id(Sports_data)
Strategy_data <- drop_id(Strategy_data)
```


```{r}
# Add Genre column to each dataset
action_data$Genre <- "Action"
adventure_data$Genre <- "Adventure"
EarlyAccess_data$Genre <- "EarlyAccess"
ExEarlyAccess_data$Genre <- "ExEarlyAccess"
Free_data$Genre <- "Free"
Indie_data$Genre <- "Indie"
MMO_data$Genre <- "MMO"
RPG_data$Genre <- "RPG"
Simulation_data$Genre <- "Simulation"
Sports_data$Genre <- "Sports"
Strategy_data$Genre <- "Strategy"

```

```{r}
head(action_data)
dim(action_data)
```

```{r}
combined_data <- bind_rows(
  action_data,
  adventure_data,
  EarlyAccess_data,
  ExEarlyAccess_data,
  Free_data,
  Indie_data,
  MMO_data,
  RPG_data,
  Simulation_data,
  Sports_data,
  Strategy_data
)

```


```{r}
# Remove duplicate rows if some games appear in multiple genres
combined_data <- unique(combined_data)

# Convert Release.date to Date type
combined_data$Release_date <- mdy(combined_data$Release.date)

# If there are NAs due to format issues, fallback to parse_date_time:
if (any(is.na(combined_data$Release_date))) {
  combined_data$Release_date <- parse_date_time(combined_data$Release.date, orders = c("mdy", "dmy", "ymd"))
}

# Extract Year
combined_data$Year <- year(combined_data$Release_date)

# Filter out rows with missing Year + is >= 2000, and specifically not 2030 as there seems to be data in 2030 for some reason
combined_data <- combined_data %>% filter(!is.na(Year) & Year >= 2000 & Year != 2030)

# Write combined data to CSV
write.csv(combined_data, "SteamSpyData/Combined_SteamSpy.csv", row.names = FALSE)

# see the combined data
head(combined_data)
```


```{r}
dim(combined_data)
```


```{r}
# Count number of games released per year per Genre
release_counts <- combined_data |>
  group_by(Year, Genre) |>
  summarise(Count = n(), .groups = "drop")

# Plot stacked bar chart
ggplot(release_counts, aes(x = factor(Year), y = Count, fill = Genre)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of Games Released by Genre Over Time",
       x = "Year",
       y = "Number of Games Released",
       fill = "Genre") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

2000 to 2010, number of games per year was low

2010 start increasing

2015 onwards, growth accelerates sharply


```{r}

# Filter data for years 2000 to 2010 inclusive
data_2000_2010 <- combined_data %>%
  filter(Year >= 2000 & Year <= 2010)

# Count number of games released per year and genre in this period
release_counts_2000_2010 <- data_2000_2010 %>%
  group_by(Year, Genre) %>%
  summarise(Count = n(), .groups = "drop")

# Plot the stacked bar chart for 2000-2010
ggplot(release_counts_2000_2010, aes(x = factor(Year), y = Count, fill = Genre)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of Games Released by Genre (2000-2010)",
       x = "Year",
       y = "Number of Games Released",
       fill = "Genre") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```





```{r}
release_counts_pct <- release_counts_2000_2010 %>%
  group_by(Year) %>%
  mutate(Percent = Count / sum(Count) * 100) %>%
  ungroup()



ggplot(release_counts_pct, aes(x = factor(Year), y = Count, fill = Genre)) +
  geom_bar(stat = "identity", position = "fill") +
  geom_text(aes(label = paste0(round(Percent, 1), "%")),
            position = position_fill(vjust = 0.5), size = 2.5, color = "white") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Proportion of Games Released by Genre (2000-2010)",
    x = "Year",
    y = "Percentage of Games Released",
    fill = "Genre"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Action and Strategy genres make up for a large share of releases every year, they were the trend from 2000 to 2010.

Indie games slowly rising in proportion, and surpass strategy in 2008 and above.

```{r}

# Filter data for 2010 and later
data_2010_current <- combined_data %>%
  filter(Year >= 2010)

# Count releases per year and genre
release_counts_2010_current <- data_2010_current %>%
  group_by(Year, Genre) %>%
  summarise(Count = n(), .groups = "drop")

# Plot stacked bar chart (absolute counts)
ggplot(release_counts_2010_current, aes(x = factor(Year), y = Count, fill = Genre)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Number of Games Released by Genre (2010 - Current)",
    x = "Year",
    y = "Number of Games Released",
    fill = "Genre"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


```{r}

release_counts_2010_plus <- combined_data %>%
  filter(Year >= 2010) %>%
  group_by(Year, Genre) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(Year) %>%
  mutate(Percent = Count / sum(Count) * 100) %>%
  ungroup()

ggplot(release_counts_2010_plus, aes(x = factor(Year), y = Count, fill = Genre)) +
  geom_bar(stat = "identity", position = "fill", width = 0.7) +  # narrower bars for spacing
  geom_text(aes(label = ifelse(Percent > 3, paste0(round(Percent, 1), "%"), "")),
            position = position_fill(vjust = 0.5), size = 2.5, color = "black") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Proportion of Games Released by Genre (2010 - Current)",
    x = "Year",
    y = "Percentage of Games Released",
    fill = "Genre"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Indie game increased in popularity and even surpass the Action genre in 2011 onwards, stragtegy genres appeared to have a less proportion.

Early and ex early access games showed an increase in release from 2014, this could indicate players believe early acces game releases are important

RPG and simulation genres mostly have no big changes throughout the entire 20 years.



So, why did indie rise ?

We will talk about 2 important game making tools: unreal engine and unity.

Unity was released for free in 2009.  Before that game engines were costly like a couple hundreds or thousands of dollars, which not everyone except AAA studios are willing to spend to develop games.  Then unity released for free which is an Indie tipping point.

Unreal engine had been out for quite a while as well, but in 2014, epic games released unreal engine 4 which is the first version of unreal engine to be free to download.  This is also a tipping point.

Both tools exist and surged the indie game development process, and we will see if it is highly correlated with the increase of indie games.

```{r}

combined_data$unity_free <- ifelse(combined_data$Year >= 2009, 1, 0)
combined_data$UE_free <- ifelse(combined_data$Year >= 2014, 1, 0)

# View the updated dataframe
head(combined_data)
```



```{r}

# Calculate yearly indie game proportions
yearly_data <- combined_data %>%
  group_by(Year) %>%
  summarise(
    indie_count = sum(Genre == "Indie", na.rm = TRUE),
    total_games = n(),
    indie_proportion = indie_count / total_games,
    unity_free = first(unity_free),  # Already 0/1 by year
    UE_free = first(UE_free)
  )

# View the processed data
yearly_data

```



```{r}

ggplot(yearly_data, aes(x = Year)) +
  geom_line(aes(y = indie_proportion * 100), color = "red", linewidth = 1.5) +
  geom_vline(xintercept = 2009, linetype = "dashed", color = "blue") +
  geom_vline(xintercept = 2014, linetype = "dashed", color = "green") +
  annotate("text", x = 2009, y = max(yearly_data$indie_proportion * 100), 
           label = "Unity Free (2009)", hjust = -0.1, color = "blue") +
  annotate("text", x = 2014, y = max(yearly_data$indie_proportion * 100), 
           label = "Unreal Free (2014)", hjust = -0.1, color = "green") +
  labs(title = "Indie Game Proportion vs. Engine Availability",
       y = "Indie Games (% of Total Releases)",
       x = "Year") +
  theme_minimal()


```


```{r}
# Calculate correlations between indie proportion and engine availability

cat("Unity Free Correlation:", cor(yearly_data$indie_proportion, yearly_data$unity_free), "\n")


cat("UE Free Correlation:",cor(yearly_data$indie_proportion, yearly_data$UE_free))
```


```{r}
ggplot(yearly_data, aes(x = Year)) +
  geom_line(aes(y = indie_count), color = "red", linewidth = 1.5, alpha = 0.5) +
  geom_point(aes(y = indie_count), color = "red", size = 3) +  # Adds points for each year
  geom_vline(xintercept = 2009, linetype = "dashed", color = "blue") +
  geom_vline(xintercept = 2014, linetype = "dashed", color = "green") +
  annotate("text", x = 2009, y = max(yearly_data$indie_count), 
           label = "Unity Free (2009)", hjust = -0.1, color = "blue") +
  annotate("text", x = 2014, y = max(yearly_data$indie_count), 
           label = "Unreal Free (2014)", hjust = -0.1, color = "green") +
  labs(title = "Indie Game Releases vs. Engine Availability",
       y = "Number of Indie Game Releases",
       x = "Year") +
  theme_minimal()
```

```{r}
# Correlation between indie count and Unity being free
cat("Unity Free Correlation (Count):", cor(yearly_data$indie_count, yearly_data$unity_free), "\n")

# Correlation between indie count and Unreal being free
cat("UE Free Correlation (Count):", cor(yearly_data$indie_count, yearly_data$UE_free), "\n")
```


Q1: Why have game genres and themes changed over the past 20 years?
A1: Expensive game engines free to public, everyone can be game devs + pour the heart, passionate -> quantity + quality 
The "pour their heart" part can be seen in steam's overwhelmingly positive review, 2/3 are indie games, showing they got quality as well


Q2: Are there emerging genres or themes that could be popular in the upcoming decade?
A2: Indie genre steady and significant release, Early Access genre had grown notably in recent years despite being small increases.


























#### Unused content




```{r}
ggplot(indie_per_year, aes(x = Year, y = indie_count)) +
  geom_point(color = "steelblue", size = 2) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  # Add vertical lines for engine availability changes
  geom_vline(xintercept = 2009, linetype = "dashed", color = "blue", linewidth = 0.8) +
  geom_vline(xintercept = 2014, linetype = "dashed", color = "green", linewidth = 0.8) +
  # Add annotations
  annotate("text", x = 2009, y = max(indie_per_year$indie_count), 
           label = "Unity Free (2009)", hjust = -0.1, color = "blue", size = 3) +
  annotate("text", x = 2014, y = max(indie_per_year$indie_count), 
           label = "Unreal Free (2014)", hjust = -0.1, color = "green", size = 3) +
  labs(
    title = "Number of Indie Games Released with Engine Availability Markers",
    subtitle = "Blue dashed line = Unity became free (2009)\nGreen dashed line = Unreal Engine became free (2014)",
    x = "Year",
    y = "Number of Indie Games"
  ) +
  theme_minimal()
```



```{r}
# Prepare data: counts + percentages per year+genre
release_counts_2010_current <- combined_data %>%
  filter(Year >= 2010) %>%
  group_by(Year, Genre) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(Year) %>%
  mutate(Percent = Count / sum(Count) * 100) %>%
  ungroup()

# Plot
ggplot(release_counts_2010_current, aes(x = factor(Year), y = Count, fill = Genre)) +
  geom_bar(stat = "identity", width = 0.7) +   # narrower bars = spacing between bars
  geom_text(aes(label = ifelse(Percent > 3, paste0(round(Percent, 1), "%"), "")),  # label only if >3%
            position = position_stack(vjust = 0.5), size = 3, color = "black") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +  # small top padding
  labs(
    title = "Number of Games Released by Genre (2010 - Current)",
    x = "Year",
    y = "Number of Games Released",
    fill = "Genre"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major.x = element_blank()  # optional: remove vertical grid lines for clarity
  )
```



```{r}



```



```{r}









```



```{r}









```



```{r}









```



```{r}









```



```{r}









```



```{r}









```



```{r}









```







